Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus (Extended Version)

### 2025-04-21 paper/paper.tex
more tweaks, one in intro and one in section 2

\ifarxiv\else {Proofs and syntactic forms elided from this paper are presented
    in full in the extended version of this paper \cite{schaefer2025intrinsicverificationparsersformal}.}\fi

@@

\paragraph{Finite Grammars}
First consider finite grammars --- those built from base types via disjunctions and
concatenations. The base types comprise characters drawn from the alphabet, the
empty string, and the empty grammar.
For each character $a$ in the alphabet we have a type $\literal a$ which
has a single parse tree for the string
\stringquote{a} and no parse trees at any other strings. The grammar $\I$ has a single
parse tree for the empty string $\epsilon = \stringquote{}$ and no parses for any other strings.
The final base type, the empty grammar $0$, has no parses for any string. We
use type-theoretic syntax to represent disjunction $\oplus$ and concatenation
$\otimes$ of
grammars. Over an input string $\w$, a parse of the disjunction $\A \oplus \B$ is either
a parse of $\A$ over the string $\w$ or a
parse of $\B$ over the string $\w$. Similarly, a parse of $\A \otimes \B$ for
$\w$ is a splitting of $w$ into two strings $\w_{\A}$ and $\w_{\B}$ with
parses for $\A$
and $\B$, respectively.

A grammar $A$ derives the word $w$ if there exists a derivation
$\internalize{w} \vdash e : A$,
where $\internalize \w$ is a context with one variable for each character of
$\w$. The term $e : A$ represents a \emph{parse tree} of $w$ for the grammar $A$. For
example, to define a parse tree for \stringquote{ab}, we use
the context $\internalize{\stringquote{ab}} = a : \literal a , b :
\literal b$. In \Cref{fig:fingram}, we give a lambda term and its
typing derivation to define a parse for a finite grammar.

\begin{figure}
\begin{floatlisting}
f : (*@$\uparrowcode$@*)('(*@\color{alphabetcolor}a@*)' (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)' (*@$\lto$@*) ('(*@\color{alphabetcolor}a@*)' (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)') (*@$\oplus$@*) '(*@\color{alphabetcolor}c@*)')
f (a , b) = inl (a , b)
\end{floatlisting}\begin{mathpar}
  \footnotesize
  \inferrule
  {
    \inferrule
    {
      \inferrule
      {~}
      {a : \literal a \vdash a : \literal a}
      \\
      \inferrule
      {~}
      {b : \literal b \vdash b : \literal b}
    }
    {a : \literal a , b : \literal b \vdash (a , b) : \literal a \otimes
      \literal b}
  }
  {a : \literal a , b : \literal b \vdash \texttt{f} := \inl(a , b) :
    (\literal a \otimes \literal b) \oplus \literal c}
\end{mathpar}
\caption{(\Agda) \stringquote{ab} is parsed by $(\literal a \otimes \literal b) \oplus \literal c$
}
\label{fig:fingram}
\end{figure}
For this interpretation of parse trees as terms to make sense, our
calculus cannot allow for \emph{any} of the usual structural rules of
type theory: weakening, contraction and exchange. Weakening allows
for variables to go unused, while contraction allows for the same
variable to be used twice, but in a parse tree, every character must
be accounted for exactly once. That is, we want to prevent the following
erroneous derivations,
\(
{a : \literal a , b : \literal b \not \vdash a : \literal a}\) and
\({a : \literal a , b : \literal b \not \vdash (a, a) : \literal a \otimes \literal a}
\).

Finally, the ordering of characters in a string cannot be ignored while
parsing, so we omit the exchange rule because it would allow
for variables in the context to be reordered. Likewise, we prevent the following derivation,
\(
  {a : \literal a , b : \literal b \not \vdash (b , a) : \literal b \otimes \literal a}
\).

\paragraph{Regular Expressions}
Regular expressions can be encoded as types generated by base types,
$\oplus$, and $\otimes$, and the Kleene star $(\cdot)^{\ast}$.  For a
grammar $\A$, we define the Kleene star $\A^{*}$ as a particular
\emph{inductive linear type} of linear lists, as shown in
\cref{fig:kleenestarinductive}. Here $\A^{*} : \L$ means we are defining
a \emph{linear} type. $\A^{*}$ has two constructors: $\nil$, which
builds a parse of type $\A^{*}$ from nothing; and $\cons$, which
linearly consumes a parse of $\A$ and a parse of $\A^{*}$ and builds a
parse of $\A^{*}$. This linear consumption is defined by the linear
function type $\lto$. The linear function type $\A \lto \B$ defines functions that
take in parses of $\A$ as input, \emph{consume} the input, and return a parse of
$\B$ as output. The arrow, $\uparrow$, wrapping these
constructors means that the constructors are not consumed
upon usage, and so are \emph{non-linear} values themselves. That is, the
names $\nil$ and $\cons$ are function symbols that may be reused as many times
as we wish.

@@

We define $h$ by recursion on terms of type $(\A \otimes \A)^{*}$.
This recursion is expressed in the derivation tree by invoking the
elimination principle for Kleene star, written
as $\fold$. The parse transformer $h$ is more compactly presented in the pseudocode
of \cref{fig:kleeneabstractproof} by pattern matching on the input and making an
explicit recursive call in the body of its definition.

@@

\paragraph{Non-deterministic Finite Automata}
\newcommand{\s}{\texttt{s}}
\newcommand{\0}{\texttt{0}}
\newcommand{\1}{\texttt{1}}
\newcommand{\2}{\texttt{2}}
Regular expressions are a compact formalism for defining a formal grammar,
but an expression such as
$({\literal a}^{*} \otimes \literal b) \oplus \literal c$ does not give an
operationalized method for parsing. For this reason, most parsers are
implemented by
compiling a grammar to a corresponding automaton, which is readily
implemented. To implement
these algorithms in \theoryabbv, we represent automata as types in the same way we can represent regular expressions.

>>>>

\ifarxiv\else {Proofs and syntactic forms elided from this paper are presented
    in full in the extended version \cite{schaefer2025intrinsicverificationparsersformal}.}\fi

@@

\paragraph{Finite Grammars}
First consider finite grammars --- those built from base types via disjunctions and
concatenations. The base types comprise characters drawn from the alphabet, the
empty string, and the empty grammar.
For each character $a$ in the alphabet we have a type $\literal a$ which
has a single parse tree for the string
\stringquote{a} and no parse trees at any other strings. The grammar $\I$ has a single
parse tree for the empty string $\epsilon = \stringquote{}$ and no parses for any other strings.
The final base type, the empty grammar $0$, has no parses for any string. We
use type-theoretic syntax to represent disjunction $\oplus$ and concatenation
$\otimes$ of
grammars. Over an input string $\w$, a parse of the disjunction $\A \oplus \B$ is either
a parse of $\A$ over the string $\w$ or a
parse of $\B$ over the string $\w$, along with a tag \texttt{inl} or \texttt{inr} indicating which case was taken. A parse of $\A \otimes \B$ for
$\w$ is a splitting of $w$ into two strings $\w_{\A}$ and $\w_{\B}$ with
parses for $\A$
and $\B$, respectively.

A derivation of a word $w$ in a grammar $A$ is given by a term in our calculus that satisfies the typing
$\internalize{w} \vdash e : A$,
where $\internalize \w$ is a context with one variable for each character of
$\w$. The term $e : A$ represents a \emph{parse tree} of $w$ for the grammar $A$. For
example, to define a parse tree for \stringquote{ab}, we use
the context $\internalize{\stringquote{ab}} = a : \literal a , b :
\literal b$. In \Cref{fig:fingram}, we give a lambda term and its
typing derivation to define a parse for a finite grammar.

\begin{figure}
\begin{floatlisting}
f : (*@$\uparrowcode$@*)('(*@\color{alphabetcolor}a@*)' (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)' (*@$\lto$@*) ('(*@\color{alphabetcolor}a@*)' (*@$\otimes$@*) '(*@\color{alphabetcolor}b@*)') (*@$\oplus$@*) '(*@\color{alphabetcolor}c@*)')
f (a , b) = inl (a , b)
\end{floatlisting}\begin{mathpar}
  \footnotesize
  \inferrule
  {
    \inferrule
    {
      \inferrule
      {~}
      {a : \literal a \vdash a : \literal a}
      \\
      \inferrule
      {~}
      {b : \literal b \vdash b : \literal b}
    }
    {a : \literal a , b : \literal b \vdash (a , b) : \literal a \otimes
      \literal b}
  }
  {a : \literal a , b : \literal b \vdash \texttt{f} := \inl(a , b) :
    (\literal a \otimes \literal b) \oplus \literal c}
\end{mathpar}
\caption{(\Agda) \stringquote{ab} is parsed by $(\literal a \otimes \literal b) \oplus \literal c$
}
\label{fig:fingram}
\end{figure}
For this interpretation of parse trees as terms to make sense, our
calculus cannot allow for \emph{any} of the usual structural rules of
type theory: weakening, contraction and exchange. Weakening allows
for variables to go unused, would allow for a character in an input string to be ignored, yielding the erroneous parse tree \( {a : \literal a , b : \literal b \not \vdash a : \literal a} \). 
Contraction allows for the same variable to be used twice, yielding the erroneous parse
\({a : \literal a , b : \literal b \not \vdash (a, a) : \literal a \otimes \literal a}
\).
Finally, the ordering of characters in a string cannot be ignored while
parsing, so we omit the exchange rule because it would allow
for variables in the context to be reordered. This prevents the erroneous derivation,
\(
  {a : \literal a , b : \literal b \not \vdash (b , a) : \literal b \otimes \literal a}
\).

\paragraph{Regular Expressions}
Regular expressions can be encoded as types generated by base types,
$\oplus$, and $\otimes$, and the Kleene star $(\cdot)^{\ast}$.  For a
grammar $\A$, we define the Kleene star $\A^{*}$ as a particular
\emph{inductive linear type} of linear lists, as shown in
\cref{fig:kleenestarinductive}. Here $\A^{*} : \L$ means we are defining $\A^*$
to be a \emph{linear} type. $\A^{*}$ has two constructors: $\nil$, which
builds a parse of type $\A^{*}$ from nothing; and $\cons$, which
linearly consumes a parse of $\A$ and a parse of $\A^{*}$ and builds a
parse of $\A^{*}$. This linear consumption is defined by the linear
function type $\lto$. The linear function type $\A \lto \B$ defines functions that
take in parses of $\A$ as input, \emph{consume} the input, and return a parse of
$\B$ as output. The arrow, $\uparrow$, wrapping these
constructors means that the constructors are not consumed
upon usage, and so are \emph{non-linear} values themselves. That is, the
names $\nil$ and $\cons$ are function symbols that may be reused as many times
as we wish.

@@

We define $h$ by recursion on terms of type $(\A \otimes \A)^{*}$.
This recursion is expressed in the derivation tree by invoking the
elimination principle for Kleene star, written
as $\fold$. The parse transformer $h$ is more intuitively presented in the pseudocode
of \cref{fig:kleeneabstractproof} by pattern matching on the input and making an
explicit recursive call in the body of its definition.

@@

\paragraph{Non-deterministic Finite Automata}
\newcommand{\s}{\texttt{s}}
\newcommand{\0}{\texttt{0}}
\newcommand{\1}{\texttt{1}}
\newcommand{\2}{\texttt{2}}
Regular expressions are a compact formalism for defining a formal grammar,
but an expression such as
$({\literal a}^{*} \otimes \literal b) \oplus \literal c$ does not give an
operationalized method for parsing. For this reason, most parsers are
implemented by
compiling a grammar to a corresponding automaton, which is readily
implemented. To implement
these algorithms in \theoryabbv, we represent automata as types, just as we did with regular expressions.
